Implementing Equational Constraints\\ in a Functional Language

### 2011-06-30 papers/wlp2011/paper.tex

\newcommand{\code}[1]{\mbox{\small\texttt{#1}}}
\newcommand{\ccode}[1]{``\code{#1}''}
\newcommand{\bs}{\char92\xspace} % backslash
\newcommand{\us}{\char95\xspace} % underscore
\newcommand{\dollar}{\code{\char36}} % dollar in programs
\newcommand{\Choice}[2]{#1 \mathop{?} #2}
\newcommand{\funset}{\ensuremath{_{\cal S}}}

@@

The condition $c$ in rule (\ref{rule}) is typically a
conjunction of \emph{equational constraints}
of the form \code{$e_1$\,=:=\,$e_2$}.
Such a constraint is satisfiable if
both sides $e_1$ and $e_2$ are evaluable to unifiable data terms.
For instance, if the symbol ``\code{++}'' denotes the usual list
concatenation operation, we can define an operation \code{last}
that computes the last element \code{e} of a non-empty list \code{xs}
as follows:
%
\begin{curry}
  last xs | ys++[e]=:=xs = e   where ys,e free  
\end{curry}
%
As in Haskell, the rules defining most functions are
\emph{constructor-based} \cite{ODonnell85}, in (\ref{rule}) $t_1
\ldots t_n$ are made of variables and/or data constructor symbols
only.  However, Curry also allows \emph{functional pattern}
\cite{AntoyHanus05LOPSTR}, i.e., $t_i$ might contain calls to
defined operations.  For instance, we can define \code{last} also
as:
%
\begin{curry}
  last (xs++[e]) = e 
\end{curry}
%
Here, the functional pattern \code{(xs++[e])}
states that \code{last} is reducible
to \code{e} provided that the argument can be matched against
some value of \code{(xs++[e])} where \code{xs} and \code{e} are
free variables.
By instantiating \code{xs} to arbitrary lists, the value
of \code{(xs++[e])} is any list having \code{e} as its last element.
Functional patterns are
a powerful feature to express arbitrary selections in term structures.
For instance, functional patterns supports a straightforward
processing of XML data with incompletely specified
or evolving formats \cite{Hanus11ICLP}.

@@

In order to extract values from a \code{Choice} structure,
we have to traverse it, compute all possible choices
but consider the choice identifiers to make consistent (left/right) choices.
The latter requirement can be implemented by storing
the choices already made during the traversal.
For this purpose, we use the type
\begin{haskell}
  data Choice = NoChoice | ChooseLeft | ChooseRight
\end{haskell}
where \code{NoChoice} represents the fact that a choice has not yet been made.
Furthermore, we assume operations to lookup the current choice
for a given identifier or change its choice (depending on the implementation
of \code{IDSupply}, KiCS2 supports various implementations
based on memory cells or finite maps):
\begin{haskell}
  lookupChoice :: ID -> IO Choice
  setChoice    :: ID -> Choice -> IO ()
\end{haskell}
%
Now we can print all values contained in a choice structure
in a depth-first manner by the following operation:\footnote{%
Note that this code has been simplified for readability
since the type system of Haskell does not
allow this direct definition.}
\label{sec:printValsDFS}
\begin{haskell}
  printValsDFS :: a -> IO ()
  printValsDFS Fail             = return ()
  printValsDFS (Choice i x1 x2) = lookupChoice i >>= choose
  printValsDFS v                = print v
   where
    choose ChooseLeft  = printValsDFS (try x1)
    choose ChooseRight = printValsDFS (try x2)
    choose NoChoice    = do newChoice ChooseLeft  x1
                            newChoice ChooseRight x2

    newChoice ch x = do setChoice i ch
                        printValsDFS (try x)
                        setChoice i NoChoice
\end{haskell}
This operation prints ignores failures and prints values
that are not a \code{Choice}. For a \code{Choice},
it checks whether a choice for
this identifier has already been made (note that the initial value
for all identifiers is \code{NoChoice}).
If a choice has been made, it follows this choice.
Otherwise, the left choice is made and stored. After printing
all the values w.r.t.\ this choice, the choice is undone (like in backtracking)
and the right choice is made and stored.

@@

As mentioned in the introduction,
our compiler translates occurrences of logic variables
into generators. For instance, each Boolean logic variable
is replaced by a generator defined like \code{aBool}.
Since such generators are standard
non-deterministic operations, they are translated like any other operation.
However, if equational constraints should be solved,
there are methods which are more efficient than generating all values.
These methods and their implementation are discussed in the
next section.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Unification}
\label{sec:Unification}

Motivate: discuss == vs. =:=

infinite search space for variable unification

distinguish free variables from choices

extend unification

add binding constraints

@@

@@

@@

% LocalWords:  Curry KiCS

>>>>

\newcommand{\code}[1]{\mbox{\small\texttt{#1}}}
\newcommand{\ccode}[1]{``\code{#1}''}
\newcommand{\bs}{\char92\xspace} % backslash
\newcommand{\us}{\char95\xspace} % underscore
\newcommand{\dollar}{\code{\char36}} % dollar in programs
\newcommand{\Choice}[2]{#1 \mathop{?} #2}
\newcommand{\todo}[1]{\fbox{\sc To do: #1}}

@@

The condition $c$ in rule (\ref{rule}) is typically a
conjunction of \emph{equational constraints}
of the form \code{$e_1$\,=:=\,$e_2$}.
Such a constraint is satisfiable if
both sides $e_1$ and $e_2$ are evaluable to unifiable data terms.
For instance, if the symbol ``\code{++}'' denotes the usual list
concatenation operation, we can define an operation \code{last}
that computes the last element \code{e} of a non-empty list \code{xs}
as follows:
%
\label{ex:last}
\begin{curry}
  last xs | ys++[e]=:=xs = e   where ys,e free  
\end{curry}
%
As in Haskell, the rules defining most functions are
\emph{constructor-based} \cite{ODonnell85}, in (\ref{rule}) $t_1
\ldots t_n$ are made of variables and/or data constructor symbols
only.  However, Curry also allows \emph{functional pattern}
\cite{AntoyHanus05LOPSTR}, i.e., $t_i$ might contain calls to
defined operations.  For instance, we can define \code{last} also
as:
%
\begin{curry}
  last (xs++[e]) = e 
\end{curry}
%
Here, the functional pattern \code{(xs++[e])}
states that \code{last} is reducible
to \code{e} provided that the argument can be matched against
some value of \code{(xs++[e])} where \code{xs} and \code{e} are
free variables.
By instantiating \code{xs} to arbitrary lists, the value
of \code{(xs++[e])} is any list having \code{e} as its last element.
Functional patterns are
a powerful feature to express arbitrary selections in term structures.
For instance, functional patterns supports a straightforward
processing of XML data with incompletely specified
or evolving formats \cite{Hanus11ICLP}.

@@

In order to extract values from a \code{Choice} structure,
we have to traverse it, compute all possible choices
but consider the choice identifiers to make consistent (left/right)
decisions.
The latter requirement can be implemented by storing
the decision already made for some choices during the traversal.
For this purpose, we use the type
\begin{haskell}
  data Decision = NoDecision | ChooseLeft | ChooseRight
\end{haskell}
where \code{NoDecision} represents the fact that the value of a choice
is yet undecided.
Furthermore, we assume operations to lookup the current decision
for a given identifier or change its decision (depending on the implementation
of \code{IDSupply}, KiCS2 supports various implementations
based on memory cells or finite maps):
\begin{haskell}
  lookupChoice :: ID -> IO Decision
  setChoice    :: ID -> Decision -> IO ()
\end{haskell}
%
Now we can print all values contained in a choice structure
in a depth-first manner by the following operation:\footnote{%
Note that this code has been simplified for readability
since the type system of Haskell does not
allow this direct definition.}
\label{sec:printValsDFS}
\begin{haskell}
  printValsDFS :: a -> IO ()
  printValsDFS Fail             = return ()
  printValsDFS (Choice i x1 x2) = lookupChoice i >>= choose
  printValsDFS v                = print v
   where
    choose ChooseLeft  = printValsDFS (try x1)
    choose ChooseRight = printValsDFS (try x2)
    choose NoDecision  = do newDecision ChooseLeft  x1
                            newDecision ChooseRight x2

    newDecision ch x = do setChoice i ch
                          printValsDFS (try x)
                          setChoice i NoDecision
\end{haskell}
This operation prints ignores failures and prints values
that are not a \code{Choice}. For a \code{Choice},
it checks whether a decision for
this identifier has already been made (note that the initial value
for all identifiers is \code{NoDecision}).
If a decision has been made for this choice, it follows this decision.
Otherwise, the left alternative is used and this decision is stored.
After printing
all the values w.r.t.\ this decision,
the decision is undone (like in backtracking)
and the right alternative is used and stored.

@@

As mentioned in the introduction,
KiCS2 translates occurrences of logic variables
into generators. For instance, consider the operation \code{not}
defined by the rules:
\begin{curry}
  not True  = False
  not False = True
\end{curry}
Then the expression \ccode{not x}, where \code{x} is a logic variable,
is translated into \ccode{not aBool} (we omit the \code{IDSupply} arguments
here). The latter expression is evaluated by reducing the
argument \code{aBool} to a choice between \code{True} or \code{False}
followed by applying \code{not} to this choice.
This is similar to a narrowing step on \ccode{not x}
that instantiates the variable \code{x} to \code{True} or \code{False}.
Actually, it has been shown in \cite{AntoyHanus06ICLP}
that narrowing expressions with logic variables
is equivalent to rewriting with generators.
Since such generators are standard
non-deterministic operations, they are translated like any other operation
so that one does not need additional run-time support for them.
However, if equational constraints should be solved,
there are methods which are more efficient than generating all values.
These methods and their implementation are discussed in the
next section.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Equational Constraints and Unification}
\label{sec:Unification}

As known from logic programming, predicates or constraints
are important to restrict the set of intended values
in a non-deterministic computation.
Apart from user-defined predicates,
equational constraints of the form \code{$e_1$\,=:=\,$e_2$}
are the most important kind of constraints.
We have already seen in the first rule defining operation
\code{last} in Section~\ref{ex:last} a typical application
of an equational constraint.

Due to the presence of non-terminating operations
and infinite data structures,
\ccode{=:=} is interpreted as the \emph{strict equality} on terms
\cite{GiovannettiLeviMoisoPalamidessi91},
i.e., the equation \code{$e_1$\,=:=\,$e_2$} is satisfied iff
$e_1$ and $e_2$ are reducible to the same ground constructor term.
In particular, expressions that do not have a value
are not equal w.r.t.\ \ccode{=:=}, e.g.,
the equational constraint \ccode{head [] =:= head []}
is not satisfiable.\footnote{From now on, we use the
standard notation for lists, i.e., \code{[]} denotes the empty list
and \code{(x:xs)} denotes a list with head element \code{x}
and tail \code{xs}.}

Due to this constructive definition, \ccode{=:=}
can be considered as a binary function defined by the
following rules (here we show only the rules for the Boolean
and list types, where \code{Success} denotes the only constructor
of the type \code{Success} of constraints):
\begin{curry}
  True   =:= True    =  Success
  False  =:= False   =  Success
  []     =:= []      =  Success
  (x:xs) =:= (y:ys)  =  x =:= y & xs =:= ys

  Success & c  =  c
\end{curry}
%
If we translate these operations into Haskell by the scheme
shown in Section~\ref{sec:Compilation},
the following rules are added to these rules
in order to propagate choices and failures:
\begin{haskell}
  Choice i l r =:= y             =  Choice i (l =:= y) (r =:= y)
  x            =:= Choice i l r  =  Choice i (x =:= l) (x =:= r)
  _            =:= _             =  Fail

  Choice i l r & c  =  Choice i (l & c) (r & c)
  _            & _  =  Fail

\end{haskell}
%
Although this is a correct implementation of equational constraints,
it might lead to an unnecessarily large search space
when it is applied to generators representing logic variables.
For instance, consider the following generator for Boolean lists:
\begin{curry}
  aBList = [] ? (aBool : aBList)
\end{curry}
This is translated into Haskell as follows:
\begin{haskell}
  aBList :: IDSupply -> [Bool]
  aBList s = Choice (thisID s) [] (aBool (leftSupply s)
                                   : aBList (rightSupply s))
\end{haskell}
Now consider the equational constraint \ccode{x =:= [True]}.
If the logic variable \code{x} is replaced by \code{aBList},
the translated expression \ccode{aBList s =:= [True]}
creates a search space when evaluating its first argument,
although there is no search required since there is only one possible
binding for \code{x}.
Furthermore and even worse, there is an infinite search space
when unifying two logic variables. For instance,
the expression \ccode{xs=:=ys \& xs++ys=:=[True]}
results in an infinite search space when the logic variables
\code{xs} and \code{ys} are replaced by generators.

To avoid these problems, we have to implement the idea
of the well-known unification principle \cite{Robinson65}.
Instead of enumerating all values for logic variables
occurring in an equational constraint,
we \emph{bind} them to another variable or term.
Since we compile into a purely functional language,
the binding cannot be performed by some side effect.
Instead, we add binding constraints to the computed results.

To implement this idea, we have to distinguish free variables
from ``standard choices'' (corresponding to occurrences of \ccode{?})
in the target code. For this purpose, we refine the definition
of the type \code{ID} as follows:\footnote{For the sake
of simplicity, we consider in the following the
implementation of \code{IDSupply} as unbounded integers.}
\begin{haskell}
  data ID = ID Integer | FreeID Integer
\end{haskell}
The new constructor \code{FreeID} marks a choice corresponding to
a free variable, e.g., the generator for Boolean variables is
redefined as
\begin{haskell}
  aBool s = Choice (FreeID (thisID s)) True False
\end{haskell}
%
If an operation is applied to a free variable and requires its value,
the free variable is transformed into a standard choice.
For this purpose, we define a simple operation to perform
this transformation:
\begin{haskell}
  narrow :: ID -> ID
  narrow (FreeID i) = ID i
  narrow x          = x
\end{haskell}
Furthermore, we use this operation in narrowing steps,
i.e., in all rules operating on \code{Choice} constructors.
For instance, in the implementation of the operation \code{not}
we replace the rule
\begin{haskell}
  not (Choice i x1 x2) s = Choice i (not x1 s) (not x2 s)
\end{haskell}
by the rule
\begin{haskell}
  not (Choice i x1 x2) s = Choice (narrow i) (not x1 s) (not x2 s)
\end{haskell}
With this small change, the run-time system can distinguish
free variables (choices with an identifier of the form \code{(FreeID$\ldots$)})
from other values. This is quite useful at the top-level
to print values containing logic variables in a specific form
rather than enumerating all values, similarly to logic programming systems.
For instance, KiCS2 evaluates the application of \code{head}
to an unknown list as follows:
\begin{haskell}
  Prelude> head xs where xs free
  {xs = (_x2:_x3)} _x2
\end{haskell}
Thus, free variables are prefixed by \code{\us{}x}.

As mentioned above, the consideration of free variables
is relevant in equational constraints where binding constraints are
generated. For this purpose, we introduce a type to represent
a \emph{binding constraints} as pair of a choice identifier
and a decision for this choice:
\begin{haskell}
  data Constraint = ID :=: Decision
\end{haskell}
Furthermore, we extend each data type by the possibility to
add some constraints:
\begin{haskell}
  data Bool   = $\ldots$ | Guard [Constraint] Bool
  data List a = $\ldots$ | Guard [Constraint] (List a)
\end{haskell}
Thus, \code{(Guard $cs$ $v$)} represents a \code{constrained value},
i.e., the value $v$ is only valid if the constraints $cs$ are
consistent. The consistency is checked when a value is extracted
from a choice structure, e.g., by the operation \code{printValsDFS},
whereas the binding constraints are created by
equational constraints: if a free variable
should be bound to a constructor, we create the same
decision as it would be done by the (successful) branch
of the generator. In case of the Boolean values,
this can be implemented by the following additional rules
for \ccode{=:=}:
\begin{haskell}
  Choice (FreeID i) _ _ =:= True   =  Guard [i :=: ChooseLeft ] Success
  Choice (FreeID i) _ _ =:= False  =  Guard [i :=: ChooseRight] Success
\end{haskell}
Hence, the binding of a variable to some known value
is implemented as a binding constraint for the choice identifier
for this variable. However, if we want to bind a variable
to another variable, we cannot store a concrete decision.
Instead, we store the information that the decisions for
both variables, when they are made to extract values,
must be identical. For this purpose we extend the decision
type to cover this information:
\begin{haskell}
  data Decision = $\ldots$ | BindTo ID
\end{haskell}
Furthermore, we add the rule that an equational constraint
between two variables yields a binding for these variables:
\begin{haskell}
  Choice (FreeID i) _ _ =:= Choice (FreeID j) _ _
    =  Guard [i :=: BindTo j] Success
\end{haskell}
%
As mentioned above, the checking of constraints
for consistency is done when values are extracted.
For this purpose, we extend the definition of the corresponding
operations by calling a constraint solver for the constraints.
For instance, the definition of \code{printValsDFS} is extended by
a rule handling constrained values:
%
\begin{haskell}
  $\ldots$
  printValsDFS (Guard constraints x) = solve constraints (printValsDFS x)
  $\ldots$
\end{haskell}
The operation \code{solve} checks the consistency of the constraints
given in the first arguments,
stores further decisions made by the constraints, e.g.,
if variables are bound together,
and evaluates its second argument.
Furthermore, the operations \code{lookupChoice} and \code{setChoice}
are extended to deal with decisions of type \code{BindTo},
i.e., they follow chains of variable bindings.

Due to the extensions described in this section,
we obtain the implementation of a unification procedure
that integrates the demand-driven evaluation of operations
occurring in the arguments.
Benchmarks evaluating the efficiency of this approach are
shown in Section~\ref{sec:Benchmarks}.

\todo{Mention accumulation of constraints in \&?}

@@

also short comparison of =:= with PAKCS and MCC

@@

Future work: adding other constraint systems

@@

% LocalWords:  Curry KiCS equational
